---
title: "Bayesian reasoning"
---

## Bayes at work

```{r}
library(tidyverse)
library(patchwork)
```


Posterior Shape: $\alpha' = \alpha + \sum y_i$
Posterior Rate: $\beta' = \beta + n$

```{r}
# parameter priors
alpha.prior <- 1.5  # shape
beta.prior <- 1.5 # rate
mean( rgamma(1e6, shape=alpha.prior, rate=beta.prior))

# data
sum_y <- 100
n  <- 10

# naive parameters based on data only
alpha.lik <- sum_y
beta.lik <- n

mean( rgamma(1e6, shape=alpha.lik, rate=beta.lik))

# parameter posteriors
alpha.post <- alpha + sum_y
beta.post <- beta + n

mean( rgamma(1e6, shape=alpha.post, rate=beta.post))

curve(dgamma(x, shape=alpha.prior, rate=beta.prior), 0, 15, col="darkgrey", lty=3, lwd=2)
curve(dgamma(x, shape=alpha.lik, rate=beta.lik), 0, 15, add=TRUE, col="red", lty=2)
curve(dgamma(x, shape=alpha.post, rate=beta.post), 0, 15, add=TRUE)
legend("topright", legend=c("prior belief", "data (naive)", "updated belief"))
```

## curve() in  ggplot
In brief,

`curve(expr,from, to) == ggplot() + geom_fun(fun, xlim)`

Simplest
```{r}
# Base R
curve(sin(x), from = 0, to = 2*pi)

# ggplot2
library(ggplot2)
ggplot() +
  geom_function(fun = sin, xlim = c(0, 2*pi))
```

One step more abstract: `ggplot() + geom_function(f, xlim=c(min, max))`
```{r}
f <- function(x) x^2 - 3*x + 1

# Base R
curve(f, from = -5, to = 5)

# ggplot2
ggplot() +
  geom_function(fun = f, xlim = c(-5, 5))
```

Multiple curves
```{r}
ggplot() +
  geom_function(aes(color = "sin"), fun = sin, xlim = c(0, 2*pi)) +
  geom_function(aes(color = "cos"), fun = cos, xlim = c(0, 2*pi)) +
  labs(color = "function")

```

curve plus data
```{r}
mydata <- tibble(
  x = -5:5,
  y = f(x)
)
ggplot(mydata, aes(x, y)) +
  geom_point() +
  geom_function(fun = f, color = "steelblue")
```

A function with extra parameters (`curve()`'s ... becomes `args = list(...)`)
```{r}
g <- function(x, k) k * x^2

ggplot() +
  geom_function(fun = g, args = list(k = 2), xlim = c(-3, 3))
```

## now dgamma in ggplot

Posterior Shape: $\alpha' = \alpha + \sum y_i$
Posterior Rate: $\beta' = \beta + n$

```{r}
# parameter priors
alpha.prior <- 2.2  # shape
beta.prior <- 1.1 # rate
mean( rgamma(1e6, shape=alpha.prior, rate=beta.prior))

# data
sum_y <- 100
n  <- 10


# naive parameters based on data only
alpha.lik <- sum_y
beta.lik <- n

set.seed(1)
data <- tibble(bug_counts = rpois(n, lambda=alpha.lik/beta.lik))

mean( rgamma(1e6, shape=alpha.lik, rate=beta.lik))

# parameter posteriors
alpha.post <- alpha + sum_y
beta.post <- beta + n

mean( rgamma(1e6, shape=alpha.post, rate=beta.post))

xmin <- 0
xmax <- 15
ymin <- 0; ymax <- 0.8

data_hist <- ggplot(data=data, aes(x=bug_counts)) + geom_histogram() +
  theme_bw() +
  labs(subtitle="Raw data")

prior_distr <- ggplot() +
  geom_function(fun=dgamma, args=list(shape = alpha.prior, rate = beta.prior),
                xlim=c(xmin, xmax), 
                n=1001, lty="dotted", color="#6FA8DC", linewidth=1) + 
  coord_cartesian(ylim=c(ymin, ymax)) + 
  labs(x="Expected average bug count", y="Probably density", subtitle="Prior understanding") +
  theme_bw() 

data_only_distr <- ggplot() +
  geom_function(fun=dgamma, args=list(shape = alpha.lik, rate = beta.lik),
                xlim=c(xmin, xmax), 
                n=1001, lty="dashed", color="#E59866", linewidth=1) +
  coord_cartesian(ylim=c(ymin, ymax)) +
  labs(x="Expected average bug count", y="Probably density", subtitle="Likelihood of data") +
  theme_bw() 

all_distr <- ggplot() +
  geom_function(aes(color = "Prior", linetype = "Prior"),
                fun = dgamma, args = list(shape = alpha.prior, rate = beta.prior),
                xlim = c(xmin, xmax), n = 1001, linewidth = 1) +
  geom_function(aes(color = "Likelihood", linetype = "Likelihood"),
                fun = dgamma, args = list(shape = alpha.lik, rate = beta.lik),
                xlim = c(xmin, xmax), n = 1001, linewidth = 1) +
  geom_function(aes(color = "Posterior", linetype = "Posterior"),
                fun = dgamma, args = list(shape = alpha.post, rate = beta.post),
                xlim = c(xmin, xmax), n = 1001, linewidth = 1) +
  coord_cartesian(ylim = c(ymin, ymax)) +
  labs(x="Expected average bug count", y="Probably density", subtitle="Updated understanding") +
  scale_color_manual(name = "Distribution",
                    breaks = c("Prior", "Likelihood", "Posterior"),
                    values = c(Prior = "#6FA8DC", Likelihood = "#E59866", Posterior = "#BAA0A1")) +
scale_linetype_manual(name = "Distribution",
                       breaks = c("Prior", "Likelihood", "Posterior"),
                       values = c(Prior = "dotted", Likelihood = "dashed", Posterior = "solid")) +
    guides(color = guide_legend(ncol = 1), linetype = guide_legend(ncol = 1)) +
  theme_bw() 

design <- "
ABC
DEF
"

plot_spacer() + data_hist + guide_area() +
  prior_distr + data_only_distr + all_distr +
  plot_layout(design = design, guides = "collect")


```

