---
title: "Chapter title"
---

![caption](picture){#fig-label width=75%}

## Goals
Explanation

In this chapter, you will, 

* stuff

## Preparatory steps

1. General background lectures: Download the slides and watch the video on XX (https://drive.google.com/drive/folders/1nWLoEGuWKIWdul9Uvse5qiB1Hf8MCwUv?usp=sharing){target="_blank"} before proceeding.
1. Data: go to our [class data repository](https://drive.google.com/drive/u/1/folders/1FR2FLiW3xyD4Er52YzVTToPZuXXReGRo){target="_blank"}, and download `XX.csv`. 
   a. Make sure the file remains in the same simple format, "comma separated values" (.csv).
   a. Put `XX.csv` in your working directory, `Rwork`.
   a. Continue to make sure the file remains in the same simple format, "comma separated values" (.csv).
1. Find out what you need for the deliverables (@sec-deliverables).

## Background on XX

## Details about our data


## Your hypothesis and predictions
**The data you have available** are ...

What you know:

1. ...
1. Your data: the abundance of herbivores remaining on plants.

Questions you might consider:

* ...

Think about what you know and ponder the above questions. Then generate a hypothesis about how various conditions might affect how ants protect plants.   **Recall a scientific hypothesis is a idea about how you think the world works.** A hypothesis differs from predictions. A hypothesis is a mechanism, not an observed result.

**After you have written down your hypothesis**, write down a prediction that follows logically and that you can evaluate with the data we have (...). Once you have a prediction, **sketch a graph** of what your data would look like if your hypothesized mechanism operated in this system.



## Data analysis


First, open R using RStudio, and **set your working directory**. You can do this in two ways:

1. using the Session menu in RStudio, or 
1. with code such as this, in which you customize it to your computer:
```
setwd("C:/Users/chenjoe/BIO209W/Rwork")
```

You can confirm or **get** your *current* working directory using this:
```{r}
getwd()
```
so when I do that, you can see my working directory is the one for this book. 

Before you proceed, make sure your working directory is your `Rwork` directory.

Second, use RStudio to create a new file of the type "R script". Save it using a name that reflects this chapter, and end the file name with ".R" or ".r".

Third, add a comment as a brief description of the script, and load any R packages you know you will need.
```{r}
#| message: false
## This script is for chapter XX in Hank's Intro to Ecology primer.

## We always use...
library(tidyverse)
## or usually just ggplot2 and dplyr packages within the tidyverse.

```





### The data	

Make sure that your data file is in your `Rwork` directory. and is still just a CSV file.

**Open the data file** (.csv file) in a spreadsheet application, a text editor, or in RStudio. Do **not** save it the file the spreadsheet application because it will try to change the format. 

1. With the file open, learn what is on the first several lines. Often, the first line or two will have *metadata* or information about the data (data about the data). 
1. Below the metadata, you will see column headings, separated by commas, because the data are *CSV* or comma-separated-values.
1. Below the column headins are the data, with columns separated by commas.

Close the data file, and *do not save it.* If you save it, chances are good that the application will try to save it in a different format. 

**Import the data** or *read* the data into R. Here is code for that. You will notice that we use a function that expects a CSV file. You will also notice that we use an *argument* to tell R how many lines of metadata to skip so it reads only the data and its column headings.
```{r}
#| echo: false
d <- read.csv("data/ArthropodObservations.csv",
              skip=1)
```

```{r}
#| eval: false
## NOTE that we skip the first line because it is metadata
d <- read.csv("ArthropodObservations.csv",
              skip=1)
```

Look at your data with one of these two:
```{r}
# glimpse(d)
# or 
str(d)
```

## Display the data to evaluate your predictions

## Deliverables {#sec-deliverables}
